---
title: "tanz-randomForest"
author: "James Topor"
date: "November 3, 2017"
output: html_document
---

## Read file from github

```{r}
tanz_m <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-imputations.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Data Prep

```{r}

# convert categoricals to factors
tanz_m$basin <- factor(tanz_m$basin)
tanz_m$region <- factor(tanz_m$region)
tanz_m$region_code <- factor(tanz_m$region_code)
tanz_m$district_code <- factor(tanz_m$district_code)
tanz_m$public_meeting_new <- factor(tanz_m$public_meeting_new)

# special edit to make scheme_management usable
tanz_m$scheme_management[which(tanz_m$scheme_management == 'None')] <- 'Other'
tanz_m$scheme_management <- factor(tanz_m$scheme_management)

tanz_m$permit_new <- factor(tanz_m$permit_new)
tanz_m$extraction_type <- factor(tanz_m$extraction_type)
tanz_m$extraction_type_class <- factor(tanz_m$extraction_type_class)
tanz_m$management <- factor(tanz_m$management)

tanz_m$payment <- factor(tanz_m$payment)
tanz_m$water_quality <- factor(tanz_m$water_quality)

tanz_m$quantity <- factor(tanz_m$quantity)
tanz_m$source <- factor(tanz_m$source)

tanz_m$waterpoint_type <- factor(tanz_m$waterpoint_type)


# convert labels to integers
tanz_m$label[which(tanz_m$label == 'functional')] <- 0
tanz_m$label[which(tanz_m$label == 'functional needs repair')] <- 1
tanz_m$label[which(tanz_m$label == 'non functional')] <- 2

tanz_m$label <- factor(tanz_m$label)

library(caret)

# split the Dataset in to train and validation subsets
set.seed(3033)
intrain <- createDataPartition(y = tanz_m$label, p= 0.7, list = FALSE)
training <- tanz_m[intrain,]
testing <- tanz_m[-intrain,]

```

Data is now prepped for model building.

_____

## Build a random forest

81.13% accuracy

```{r}
library(randomForest)

# NOTE: variables were modified due to incompatibility w submit data. region
# replaces region_code; extraction_type_class replaces extraction_type; 
# scheme_management 'None' values were set to 'Other'
rf_mod <- randomForest(label ~                        
quantity + waterpoint_type + extraction_type_class + pump_age + management +
                      region + district_code + water_quality + payment +
                      basin + source + scheme_management + latitude + longitude +
                      public_meeting_new + permit_new + amount_tsh_new + 
                      population_new +
                      installer_0 +
                      installer_cen +
                      installer_ces +
                      installer_com +
                      installer_dan +
                      installer_dis +
                      installer_dwe +
                      installer_fin +
                      installer_gov +
                      installer_hes +
                      installer_kkk +
                      installer_other +
                      installer_rwe +
                      installer_tcr +
                      installer_unk +
                      installer_wor +
                      funder_dan +
                      funder_dhv +
                      funder_dis +
                      funder_dws +
                      funder_fin +
                      funder_gov +
                      funder_hes +
                      funder_kkk +
                      funder_other +
                      funder_pri +
                      funder_rws +
                      funder_tas +
                      funder_uni +
                      funder_unk +
                      funder_wat +
                      funder_wor +
                      lga_freq_bin_1 +
                      lga_freq_bin_10 +
                      lga_freq_bin_11 +
                      lga_freq_bin_2 +
                      lga_freq_bin_20 +
                      lga_freq_bin_3 +
                      lga_freq_bin_4 +
                      lga_freq_bin_5 +
                      lga_freq_bin_6 +
                      lga_freq_bin_7 +
                      lga_freq_bin_8 +
                      lga_freq_bin_9 +
                      ward_freq_bin_1 +
                      ward_freq_bin_10 +
                      ward_freq_bin_11 +
                      ward_freq_bin_12 +
                      ward_freq_bin_13 +
                      ward_freq_bin_14 +
                      ward_freq_bin_15 +
                      ward_freq_bin_16 +
                      ward_freq_bin_17 +
                      ward_freq_bin_2 +
                      ward_freq_bin_20 +
                      ward_freq_bin_3 +
                      ward_freq_bin_4 +
                      ward_freq_bin_5 +
                      ward_freq_bin_6 +
                      ward_freq_bin_7 +
                      ward_freq_bin_8 +
                      ward_freq_bin_9 +

                      subvillage_freq_bin_1 +
                      subvillage_freq_bin_10 +
                      subvillage_freq_bin_11 +
                      subvillage_freq_bin_15 +
                      subvillage_freq_bin_2 +
                      subvillage_freq_bin_20 +
                      subvillage_freq_bin_3 +
                      subvillage_freq_bin_4 +
                      subvillage_freq_bin_5 +
                      subvillage_freq_bin_6 +
                      subvillage_freq_bin_7 +
                      subvillage_freq_bin_8 +
                      gps_height_new,

                  ntree =700,

              #   data = training)
               data = tanz_m)

# summary(rf_mod)

# remove the response variable from the validation subset
# rf_testing <- subset(testing, select = -c(label))
rf_testing <- subset(tanz_m, select = -c(label))

# make predictions for the validation
rf1 <- predict(rf_mod, rf_testing)

# confusion matrix - validation subset
# rf_tab <- table(rf1, testing$label)
rf_tab <- table(rf1, tanz_m$label)

# display: off-diag numbers are mis-classifications
confusionMatrix(rf_tab)

# memory cleanup
rm(rf1, rf_testing, rf_tab)
```

## Confusion Matrix and Statistics

   
Confusion Matrix and Statistics

   
rf1     0     1     2
  0 31467  1567  2394
  1   184  2471   144
  2   608   279 20286

Overall Statistics
                                          
               Accuracy : 0.9129          
                 95% CI : (0.9106, 0.9151)
    No Information Rate : 0.5431          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.8373          
 Mcnemar's Test P-Value : < 2.2e-16       

Statistics by Class:

                     Class: 0 Class: 1 Class: 2
Sensitivity            0.9754  0.57239   0.8888
Specificity            0.8541  0.99405   0.9757
Pos Pred Value         0.8882  0.88282   0.9581
Neg Pred Value         0.9670  0.96739   0.9336
Prevalence             0.5431  0.07268   0.3842
Detection Rate         0.5297  0.04160   0.3415
Detection Prevalence   0.5964  0.04712   0.3564
Balanced Accuracy      0.9148  0.78322   0.9323

_____

## Check variable importance:

```{r}
importance <- varImp(rf_mod)

importance <- importance[order(importance$Overall,decreasing = TRUE), ,drop = FALSE]

print(importance)

# memory cleanup
rm(rf1, rf_testing, rf_tab)
```

                       Overall
quantity           2386.462153
longitude          1280.208366
latitude           1243.581667
extraction_type    1178.682671
pump_age           1116.002314
waterpoint_type    1093.158956
payment             816.229526
region_code         681.391666
population_new      661.159021
source              578.178014
district_code       555.321904
basin               427.311991
scheme_management   384.212795
management          372.834456
amount_tsh_new      345.614271
water_quality       299.133328
funder_other        153.621007
ward_freq_bin_3     149.327702
ward_freq_bin_2     142.580870
funder_gov          135.037070
installer_other     132.631561
installer_dwe       123.834254
permit_new          121.603216
ward_freq_bin_4     110.478197
public_meeting_new  108.145648
ward_freq_bin_1     102.555506
ward_freq_bin_5      89.906952
lga_freq_bin_4       80.632662
lga_freq_bin_3       78.191339
lga_freq_bin_6       74.982971
lga_freq_bin_5       71.640857
lga_freq_bin_10      61.760992
lga_freq_bin_2       61.629666
lga_freq_bin_7       60.363336
installer_gov        58.517909
ward_freq_bin_6      57.437098
installer_cen        57.360212
funder_wor           57.045224
installer_rwe        54.030158
lga_freq_bin_9       49.896391
ward_freq_bin_7      49.652331
ward_freq_bin_8      48.296252
funder_dan           39.977137
installer_unk        37.709373
lga_freq_bin_8       37.567692
funder_hes           36.508775
lga_freq_bin_20      35.405233
installer_dis        35.347900
funder_unk           34.678247
installer_com        34.384736
funder_tas           33.928214
funder_dws           31.883431
funder_pri           29.892122
funder_uni           25.143246
ward_freq_bin_9      24.503300
funder_wat           24.246632
installer_wor        23.716591
funder_rws           22.778945
installer_hes        20.582274
funder_dis           18.413152
installer_tcr        18.133932
installer_dan        17.600180
funder_kkk           17.576963
installer_fin        15.507702
lga_freq_bin_11      14.531972
installer_kkk        14.371995
funder_fin           14.235434
lga_freq_bin_1       14.198987
funder_dhv           13.691042
ward_freq_bin_10     10.414861
ward_freq_bin_14      9.585436
ward_freq_bin_13      9.147670
installer_0           7.232043
ward_freq_bin_12      6.411819
ward_freq_bin_16      5.501362
installer_ces         5.378000
ward_freq_bin_11      3.980340
ward_freq_bin_15      3.305827
ward_freq_bin_20      2.600001
ward_freq_bin_17      2.306922

## Make predictions for submit data set

```{r}
# 
tanz_subm <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-submit-imps.csv", header = TRUE, stringsAsFactors = FALSE)

# convert categoricals to factors
tanz_subm$basin <- factor(tanz_subm$basin)
tanz_subm$region <- factor(tanz_subm$region)
tanz_subm$region_code <- factor(tanz_subm$region_code)
tanz_subm$district_code <- factor(tanz_subm$district_code)
tanz_subm$public_meeting_new <- factor(tanz_subm$public_meeting_new)
tanz_subm$scheme_management <- factor(tanz_subm$scheme_management)

tanz_subm$permit_new <- factor(tanz_subm$permit_new)
tanz_subm$extraction_type <- factor(tanz_subm$extraction_type)
tanz_subm$extraction_type_class <- factor(tanz_subm$extraction_type_class)
tanz_subm$management <- factor(tanz_subm$management)

tanz_subm$payment <- factor(tanz_subm$payment)
tanz_subm$water_quality <- factor(tanz_subm$water_quality)

tanz_subm$quantity <- factor(tanz_subm$quantity)
tanz_subm$source <- factor(tanz_subm$source)

tanz_subm$waterpoint_type <- factor(tanz_subm$waterpoint_type)


# make predictions for the validation
rfsub1 <- predict(rf_mod, tanz_subm)

# add predictions to submit dataframe
# add predictions to submit dataframe
tanz_subm$status_group <- as.character(rfsub1)

# convert integer labels to corresponding text
tanz_subm$status_group[which(tanz_subm$status_group == '0')] <- 'functional'
tanz_subm$status_group[which(tanz_subm$status_group == '1')] <- 'functional needs repair'
tanz_subm$status_group[which(tanz_subm$status_group == '2')] <- 'non functional'

# write id + status_group cols to a file for submission

rf_preds <- cbind(tanz_subm$id, tanz_subm$status_group)
colnames(rf_preds) <- c("id", "status_group")


write.csv(rf_preds, file = "C:/data/698/rf_preds.csv", row.names = FALSE)

rm(rf_preds, rfsub1)

```

