---
title: "tanz-svm.Rmd"
author: "James Topor"
date: "November 3, 2017"
output: html_document
---

## Read file from github

```{r}
tanz_m <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-imputations.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Data Prep

```{r}

# convert categoricals to factors
tanz_m$basin <- factor(tanz_m$basin)
tanz_m$region <- factor(tanz_m$region)
tanz_m$region_code <- factor(tanz_m$region_code)
tanz_m$district_code <- factor(tanz_m$district_code)
tanz_m$public_meeting_new <- factor(tanz_m$public_meeting_new)
tanz_m$scheme_management <- factor(tanz_m$scheme_management)
tanz_m$permit_new <- factor(tanz_m$permit_new)
tanz_m$extraction_type <- factor(tanz_m$extraction_type)
tanz_m$extraction_type_class <- factor(tanz_m$extraction_type_class)
tanz_m$management <- factor(tanz_m$management)

tanz_m$payment <- factor(tanz_m$payment)
tanz_m$water_quality <- factor(tanz_m$water_quality)

tanz_m$quantity <- factor(tanz_m$quantity)
tanz_m$source <- factor(tanz_m$source)

tanz_m$waterpoint_type <- factor(tanz_m$waterpoint_type)


# convert labels to integers
tanz_m$label[which(tanz_m$label == 'functional')] <- 0
tanz_m$label[which(tanz_m$label == 'functional needs repair')] <- 1
tanz_m$label[which(tanz_m$label == 'non functional')] <- 2

tanz_m$label <- factor(tanz_m$label)

library(caret)

# split the Dataset in to train and validation subsets
set.seed(3033)
intrain <- createDataPartition(y = tanz_m$label, p= 0.7, list = FALSE)
training <- tanz_m[intrain,]
testing <- tanz_m[-intrain,]

```

Data is now prepped for model building.

_____

## Build a supplort vector machine

Achieves 76.5% accuracy on validation subset

```{r}
library(kernlab)

# NOTE: variables were modified due to incompatibility w submit data. region
# replaces region_code; extraction_type_class replaces extraction_type; 
# scheme_management 'None' values were set to 'Other'

svm_mod <- ksvm(label ~                        
quantity + waterpoint_type + extraction_type_class + pump_age + management +
                      region + district_code + water_quality + payment +
                      basin + source + scheme_management + latitude + longitude +
                      public_meeting_new + permit_new + amount_tsh_new + 
                      population_new +
                      installer_0 +
                      installer_cen +
                      installer_ces +
                      installer_com +
                      installer_dan +
                      installer_dis +
                      installer_dwe +
                      installer_fin +
                      installer_gov +
                      installer_hes +
                      installer_kkk +
                      installer_other +
                      installer_rwe +
                      installer_tcr +
                      installer_unk +
                      installer_wor +
                      funder_dan +
                      funder_dhv +
                      funder_dis +
                      funder_dws +
                      funder_fin +
                      funder_gov +
                      funder_hes +
                      funder_kkk +
                      funder_other +
                      funder_pri +
                      funder_rws +
                      funder_tas +
                      funder_uni +
                      funder_unk +
                      funder_wat +
                      funder_wor +
                      lga_freq_bin_1 +
                      lga_freq_bin_10 +
                      lga_freq_bin_11 +
                      lga_freq_bin_2 +
                      lga_freq_bin_20 +
                      lga_freq_bin_3 +
                      lga_freq_bin_4 +
                      lga_freq_bin_5 +
                      lga_freq_bin_6 +
                      lga_freq_bin_7 +
                      lga_freq_bin_8 +
                      lga_freq_bin_9 +
                      ward_freq_bin_1 +
                      ward_freq_bin_10 +
                      ward_freq_bin_11 +
                      ward_freq_bin_12 +
                      ward_freq_bin_13 +
                      ward_freq_bin_14 +
                      ward_freq_bin_15 +
                      ward_freq_bin_16 +
                      ward_freq_bin_17 +
                      ward_freq_bin_2 +
                      ward_freq_bin_20 +
                      ward_freq_bin_3 +
                      ward_freq_bin_4 +
                      ward_freq_bin_5 +
                      ward_freq_bin_6 +
                      ward_freq_bin_7 +
                      ward_freq_bin_8 +
                      ward_freq_bin_9,

                 data = tanz_m)

# summary(svm_mod)

svm_testing <- subset(tanz_m, select = -c(label))
svm1 <- predict(svm_mod,svm_testing, type = "response")

# make predictions
# confusion matrix - validation subset
svm_tab <- table(svm1, tanz_m$label)
# display: off-diag numbers are mis-classifications
confusionMatrix(svm_tab)

# memory cleanup
rm(svm1, svm_testing, svm_tab)

```

## Confusion Matrix and Statistics

    
svm1    0    1    2
   0 9014  938 2353
   1   46  150   27
   2  617  207 4467

Overall Statistics
                                          
               Accuracy : 0.765           
                 95% CI : (0.7587, 0.7712)
    No Information Rate : 0.5431          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.5391          
 Mcnemar's Test P-Value : < 2.2e-16       

Statistics by Class:

                     Class: 0 Class: 1 Class: 2
Sensitivity            0.9315 0.115830   0.6524
Specificity            0.5958 0.995582   0.9249
Pos Pred Value         0.7325 0.672646   0.8443
Neg Pred Value         0.8798 0.934928   0.8100
Prevalence             0.5431 0.072675   0.3843
Detection Rate         0.5059 0.008418   0.2507
Detection Prevalence   0.6906 0.012515   0.2969
Balanced Accuracy      0.7636 0.555706   0.7887


## Check variable importance:

No way to do this for an SVM so this is set to eval = FALSE

```{r, eval = FALSE}
importance <- varImp(svm_mod)

importance <- importance[order(importance$Overall,decreasing = TRUE), ,drop = FALSE]

print(importance)

# memory cleanup
rm(svm1, svm_testing, svm_tab)
```


## Make predictions for submit data set

```{r}
# 
tanz_subm <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-submit-imps.csv", header = TRUE, stringsAsFactors = FALSE)

# convert categoricals to factors
tanz_subm$basin <- factor(tanz_subm$basin)
tanz_subm$region <- factor(tanz_subm$region)
tanz_subm$region_code <- factor(tanz_subm$region_code)
tanz_subm$district_code <- factor(tanz_subm$district_code)
tanz_subm$public_meeting_new <- factor(tanz_subm$public_meeting_new)
tanz_subm$scheme_management <- factor(tanz_subm$scheme_management)

tanz_subm$permit_new <- factor(tanz_subm$permit_new)
tanz_subm$extraction_type <- factor(tanz_subm$extraction_type)
tanz_subm$extraction_type_class <- factor(tanz_subm$extraction_type_class)
tanz_subm$management <- factor(tanz_subm$management)

tanz_subm$payment <- factor(tanz_subm$payment)
tanz_subm$water_quality <- factor(tanz_subm$water_quality)

tanz_subm$quantity <- factor(tanz_subm$quantity)
tanz_subm$source <- factor(tanz_subm$source)

tanz_subm$waterpoint_type <- factor(tanz_subm$waterpoint_type)


# make predictions for the validation
svmsub1 <- predict(svm_mod, tanz_subm)

# add predictions to submit dataframe
# add predictions to submit dataframe
tanz_subm$status_group <- as.character(svmsub1)

# convert integer labels to corresponding text
tanz_subm$status_group[which(tanz_subm$status_group == '0')] <- 'functional'
tanz_subm$status_group[which(tanz_subm$status_group == '1')] <- 'functional needs repair'
tanz_subm$status_group[which(tanz_subm$status_group == '2')] <- 'non functional'

# write id + status_group cols to a file for submission

svm_preds <- cbind(tanz_subm$id, tanz_subm$status_group)
colnames(svm_preds) <- c("id", "status_group")


write.csv(svm_preds, file = "C:/data/698/svm_preds.csv", row.names = FALSE)

rm(svm_preds, svmsub1)

```
