---
title: "tanz-bag.Rmd"
author: "James Topor"
date: "November 3, 2017"
output: html_document
editor_options: 
  chunk_output_type: console
---


# Bagging Classification and Regression Trees (BCART)


## Read file from github

```{r}
tanz_m <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-imputations.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Data Prep

```{r}

# convert categoricals to factors
tanz_m$basin <- factor(tanz_m$basin)
tanz_m$region <- factor(tanz_m$region)
tanz_m$region_code <- factor(tanz_m$region_code)
tanz_m$district_code <- factor(tanz_m$district_code)
tanz_m$public_meeting_new <- factor(tanz_m$public_meeting_new)
tanz_m$scheme_management <- factor(tanz_m$scheme_management)
tanz_m$permit_new <- factor(tanz_m$permit_new)
tanz_m$extraction_type <- factor(tanz_m$extraction_type)
tanz_m$extraction_type_class <- factor(tanz_m$extraction_type_class)
tanz_m$management <- factor(tanz_m$management)

tanz_m$payment <- factor(tanz_m$payment)
tanz_m$water_quality <- factor(tanz_m$water_quality)

tanz_m$quantity <- factor(tanz_m$quantity)
tanz_m$source <- factor(tanz_m$source)

tanz_m$waterpoint_type <- factor(tanz_m$waterpoint_type)


# convert labels to integers
tanz_m$label[which(tanz_m$label == 'functional')] <- 0
tanz_m$label[which(tanz_m$label == 'functional needs repair')] <- 1
tanz_m$label[which(tanz_m$label == 'non functional')] <- 2

tanz_m$label <- factor(tanz_m$label)

library(caret)

# split the Dataset in to train and validation subsets
set.seed(3033)
intrain <- createDataPartition(y = tanz_m$label, p= 0.7, list = FALSE)
training <- tanz_m[intrain,]
testing <- tanz_m[-intrain,]

```

Data is now prepped for model building.

_____

## Build a booststrap aggregation algorithm

Bootstrap Aggregation

Achieves 79.91% accuracy on validation subset

```{r, eval = FALSE}
library(ipred)

bag_mod <- bagging(label ~                        
quantity + waterpoint_type + extraction_type + pump_age + management +
                      region_code + district_code + water_quality + payment +
                      basin + source + scheme_management + latitude + longitude +
                      public_meeting_new + permit_new + amount_tsh_new + 
                      population_new +
                      installer_0 +
                      installer_cen +
                      installer_ces +
                      installer_com +
                      installer_dan +
                      installer_dis +
                      installer_dwe +
                      installer_fin +
                      installer_gov +
                      installer_hes +
                      installer_kkk +
                      installer_other +
                      installer_rwe +
                      installer_tcr +
                      installer_unk +
                      installer_wor +
                      funder_dan +
                      funder_dhv +
                      funder_dis +
                      funder_dws +
                      funder_fin +
                      funder_gov +
                      funder_hes +
                      funder_kkk +
                      funder_other +
                      funder_pri +
                      funder_rws +
                      funder_tas +
                      funder_uni +
                      funder_unk +
                      funder_wat +
                      funder_wor +
                      lga_freq_bin_1 +
                      lga_freq_bin_10 +
                      lga_freq_bin_11 +
                      lga_freq_bin_2 +
                      lga_freq_bin_20 +
                      lga_freq_bin_3 +
                      lga_freq_bin_4 +
                      lga_freq_bin_5 +
                      lga_freq_bin_6 +
                      lga_freq_bin_7 +
                      lga_freq_bin_8 +
                      lga_freq_bin_9 +
                      ward_freq_bin_1 +
                      ward_freq_bin_10 +
                      ward_freq_bin_11 +
                      ward_freq_bin_12 +
                      ward_freq_bin_13 +
                      ward_freq_bin_14 +
                      ward_freq_bin_15 +
                      ward_freq_bin_16 +
                      ward_freq_bin_17 +
                      ward_freq_bin_2 +
                      ward_freq_bin_20 +
                      ward_freq_bin_3 +
                      ward_freq_bin_4 +
                      ward_freq_bin_5 +
                      ward_freq_bin_6 +
                      ward_freq_bin_7 +
                      ward_freq_bin_8 +
                      ward_freq_bin_9 +
  
                    subvillage_freq_bin_1 +
                      subvillage_freq_bin_10 +
                      subvillage_freq_bin_11 +
                      subvillage_freq_bin_15 +
                      subvillage_freq_bin_2 +
                     subvillage_freq_bin_20 +
                      subvillage_freq_bin_3 +
                      subvillage_freq_bin_4 +
                      subvillage_freq_bin_5 +
                      subvillage_freq_bin_6 +
                      subvillage_freq_bin_7 +
                      subvillage_freq_bin_8 +
                   gps_height_new,

                 data = tanz_m)

# summary(bag_mod)

bag_testing <- subset(tanz_m, select = -c(label))
bag1 <- predict(bag_mod, bag_testing)

# make predictions
# confusion matrix - validation subset
bag_tab <- table(bag1, tanz_m$label)
# display: off-diag numbers are mis-classifications
confusionMatrix(bag_tab)

# calculate percentage of misclassifications
# sum(diag(svm1))/sum(svm1)

# memory cleanup
rm(bag1, bag_testing, bag_tab)
```

Confusion Matrix and Statistics

    
bag1     0     1     2
   0 32164   191   232
   1    39  4100    17
   2    56    26 22575

Overall Statistics
                                          
               Accuracy : 0.9906          
                 95% CI : (0.9897, 0.9913)
    No Information Rate : 0.5431          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.9828          
 Mcnemar's Test P-Value : < 2.2e-16       

Statistics by Class:

                     Class: 0 Class: 1 Class: 2
Sensitivity            0.9971  0.94973   0.9891
Specificity            0.9844  0.99898   0.9978
Pos Pred Value         0.9870  0.98653   0.9964
Neg Pred Value         0.9965  0.99607   0.9932
Prevalence             0.5431  0.07268   0.3842
Detection Rate         0.5415  0.06902   0.3801
Detection Prevalence   0.5486  0.06997   0.3814
Balanced Accuracy      0.9907  0.97436   0.9934

## Make predictions for submit data set

```{r}
# 
tanz_subm <- read.csv("C:/data/698/tanz-submit-imps.csv", header = TRUE, stringsAsFactors = FALSE)

# convert categoricals to factors
tanz_subm$basin <- factor(tanz_subm$basin)
tanz_subm$region <- factor(tanz_subm$region)
tanz_subm$region_code <- factor(tanz_subm$region_code)
tanz_subm$district_code <- factor(tanz_subm$district_code)
tanz_subm$public_meeting_new <- factor(tanz_subm$public_meeting_new)
tanz_subm$scheme_management <- factor(tanz_subm$scheme_management)
tanz_subm$permit_new <- factor(tanz_subm$permit_new)
tanz_subm$extraction_type <- factor(tanz_subm$extraction_type)
tanz_subm$extraction_type_class <- factor(tanz_subm$extraction_type_class)
tanz_subm$management <- factor(tanz_subm$management)

tanz_subm$payment <- factor(tanz_subm$payment)
tanz_subm$water_quality <- factor(tanz_subm$water_quality)

tanz_subm$quantity <- factor(tanz_subm$quantity)
tanz_subm$source <- factor(tanz_subm$source)

tanz_subm$waterpoint_type <- factor(tanz_subm$waterpoint_type)


# make predictions for the validation
bagsub1 <- predict(bag_mod, tanz_subm)

# add predictions to submit dataframe
tanz_subm$status_group <- as.character(bagsub1)


# convert integer labels to corresponding text
tanz_subm$status_group[which(tanz_subm$status_group == '0')] <- 'functional'
tanz_subm$status_group[which(tanz_subm$status_group == '1')] <- 'functional needs repair'
tanz_subm$status_group[which(tanz_subm$status_group == '2')] <- 'non functional'

# write id + status_group cols to a file for submission

bag_preds <- cbind(tanz_subm$id, tanz_subm$status_group)
colnames(bag_preds) <- c("id", "status_group")

write.csv(bag_preds, file = "C:/data/698/bag_preds.csv", row.names = FALSE)

rm(bag_preds, bagsub1)

```
