---
title: "tanz-knn.Rmd"
author: "James Topor"
date: "November 3, 2017"
output: html_document
---

## Read file from github

```{r}
tanz_m <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-imputations.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Data Prep

```{r}

# convert categoricals to factors
tanz_m$basin <- factor(tanz_m$basin)
tanz_m$region <- factor(tanz_m$region)
tanz_m$region_code <- factor(tanz_m$region_code)
tanz_m$district_code <- factor(tanz_m$district_code)
tanz_m$public_meeting_new <- factor(tanz_m$public_meeting_new)
tanz_m$scheme_management <- factor(tanz_m$scheme_management)
tanz_m$permit_new <- factor(tanz_m$permit_new)
tanz_m$extraction_type <- factor(tanz_m$extraction_type)
tanz_m$extraction_type_class <- factor(tanz_m$extraction_type_class)
tanz_m$management <- factor(tanz_m$management)

tanz_m$payment <- factor(tanz_m$payment)
tanz_m$water_quality <- factor(tanz_m$water_quality)

tanz_m$quantity <- factor(tanz_m$quantity)
tanz_m$source <- factor(tanz_m$source)

tanz_m$waterpoint_type <- factor(tanz_m$waterpoint_type)


# convert labels to integers
tanz_m$label[which(tanz_m$label == 'functional')] <- 0
tanz_m$label[which(tanz_m$label == 'functional needs repair')] <- 1
tanz_m$label[which(tanz_m$label == 'non functional')] <- 2

tanz_m$label <- factor(tanz_m$label)

library(caret)

# split the Dataset in to train and validation subsets
set.seed(3033)
intrain <- createDataPartition(y = tanz_m$label, p= 0.7, list = FALSE)
training <- tanz_m[intrain,]
testing <- tanz_m[-intrain,]

```

Data is now prepped for model building.

_____


## Build a K-Nearest Neighbors model using caret package

74.93% accurate


```{r}

# train a KNN model
# trctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
trctrl <- trainControl(method = "repeatedcv", number = 2, repeats = 1)
set.seed(3333)

knn_fit <- train(label ~                        
quantity + waterpoint_type + extraction_type + pump_age + management +
                      region_code + district_code + water_quality + payment +
                      basin + source + scheme_management + latitude + longitude +
                      public_meeting_new + permit_new + amount_tsh_new + 
                      population_new +
                      installer_0 +
                      installer_cen +
                      installer_ces +
                      installer_com +
                      installer_dan +
                      installer_dis +
                      installer_dwe +
                      installer_fin +
                      installer_gov +
                      installer_hes +
                      installer_kkk +
                      installer_other +
                      installer_rwe +
                      installer_tcr +
                      installer_unk +
                      installer_wor +
                      funder_dan +
                      funder_dhv +
                      funder_dis +
                      funder_dws +
                      funder_fin +
                      funder_gov +
                      funder_hes +
                      funder_kkk +
                      funder_other +
                      funder_pri +
                      funder_rws +
                      funder_tas +
                      funder_uni +
                      funder_unk +
                      funder_wat +
                      funder_wor +
                      lga_freq_bin_1 +
                      lga_freq_bin_10 +
                      lga_freq_bin_11 +
                      lga_freq_bin_2 +
                      lga_freq_bin_20 +
                      lga_freq_bin_3 +
                      lga_freq_bin_4 +
                      lga_freq_bin_5 +
                      lga_freq_bin_6 +
                      lga_freq_bin_7 +
                      lga_freq_bin_8 +
                      lga_freq_bin_9 +
                      ward_freq_bin_1 +
                      ward_freq_bin_10 +
                      ward_freq_bin_11 +
                      ward_freq_bin_12 +
                      ward_freq_bin_13 +
                      ward_freq_bin_14 +
                      ward_freq_bin_15 +
                      ward_freq_bin_16 +
                      ward_freq_bin_17 +
                      ward_freq_bin_2 +
                      ward_freq_bin_20 +
                      ward_freq_bin_3 +
                      ward_freq_bin_4 +
                      ward_freq_bin_5 +
                      ward_freq_bin_6 +
                      ward_freq_bin_7 +
                      ward_freq_bin_8 +
                      ward_freq_bin_9,

                      data = tanz_m, method = "knn",
                      trControl=trctrl,
                      preProcess = c("center", "scale"),
                      tuneLength = 10
        )

# check results
knn_fit

# THis model scores 74.93% accuracy w k = 5 found as best k value 

```

k-Nearest Neighbors 

41581 samples
   80 predictor
    3 classes: '0', '1', '2' 

Pre-processing: centered (194), scaled (194) 
Resampling: Cross-Validated (5 fold, repeated 1 times) 
Summary of sample sizes: 33265, 33265, 33265, 33264, 33265 
Resampling results across tuning parameters:

  k   Accuracy   Kappa    
   5  0.7455086  0.5250471
   7  0.7420936  0.5158868
   9  0.7372597  0.5046890
  11  0.7347825  0.4987192
  13  0.7314156  0.4910240
  15  0.7279527  0.4829173
  17  0.7238883  0.4742496
  19  0.7216517  0.4694295
  21  0.7206897  0.4667426
  23  0.7196075  0.4635517

Accuracy was used to select the optimal model using  the largest value.
The final value used for the model was k = 5.

```{r}
# get predictions
test_pred <- predict(knn_fit, newdata = tanz_m)
# test_pred

# check confusion matrix
confusionMatrix(test_pred, tanz_m$label )

# memory cleanup
rm(test_pred)
 
```

Confusion Matrix and Statistics

          Reference
Prediction    0    1    2
         0 8078  621 1837
         1  310  418  154
         2 1289  256 4856

Overall Statistics
                                          
               Accuracy : 0.7493          
                 95% CI : (0.7429, 0.7557)
    No Information Rate : 0.5431          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.5334          
 Mcnemar's Test P-Value : < 2.2e-16       

Statistics by Class:

                     Class: 0 Class: 1 Class: 2
Sensitivity            0.8348  0.32278   0.7092
Specificity            0.6981  0.97192   0.8592
Pos Pred Value         0.7667  0.47392   0.7586
Neg Pred Value         0.7804  0.94822   0.8256
Prevalence             0.5431  0.07268   0.3843
Detection Rate         0.4533  0.02346   0.2725
Detection Prevalence   0.5913  0.04950   0.3592
Balanced Accuracy      0.7664  0.64735   0.7842



## Make predictions for submit data set

```{r}
# 
tanz_subm <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-submit-imps.csv", header = TRUE, stringsAsFactors = FALSE)

# convert categoricals to factors
tanz_subm$basin <- factor(tanz_subm$basin)
tanz_subm$region <- factor(tanz_subm$region)
tanz_subm$region_code <- factor(tanz_subm$region_code)
tanz_subm$district_code <- factor(tanz_subm$district_code)
tanz_subm$public_meeting_new <- factor(tanz_subm$public_meeting_new)
tanz_subm$scheme_management <- factor(tanz_subm$scheme_management)

tanz_subm$permit_new <- factor(tanz_subm$permit_new)
tanz_subm$extraction_type <- factor(tanz_subm$extraction_type)
tanz_subm$extraction_type_class <- factor(tanz_subm$extraction_type_class)
tanz_subm$management <- factor(tanz_subm$management)

tanz_subm$payment <- factor(tanz_subm$payment)
tanz_subm$water_quality <- factor(tanz_subm$water_quality)

tanz_subm$quantity <- factor(tanz_subm$quantity)
tanz_subm$source <- factor(tanz_subm$source)

tanz_subm$waterpoint_type <- factor(tanz_subm$waterpoint_type)


# make predictions for the validation
knnsub1 <- predict(knn_fit, tanz_subm)

# add predictions to submit dataframe
# add predictions to submit dataframe
tanz_subm$status_group <- as.character(knnsub1)

# convert integer labels to corresponding text
tanz_subm$status_group[which(tanz_subm$status_group == '0')] <- 'functional'
tanz_subm$status_group[which(tanz_subm$status_group == '1')] <- 'functional needs repair'
tanz_subm$status_group[which(tanz_subm$status_group == '2')] <- 'non functional'

# write id + status_group cols to a file for submission

knn_preds <- cbind(tanz_subm$id, tanz_subm$status_group)
colnames(knn_preds) <- c("id", "status_group")


write.csv(knn_preds, file = "C:/data/698/knn_preds.csv", row.names = FALSE)

rm(knn_preds, knnsub1)

```

