---
title: "tanz-gbm"
author: "James Topor"
date: "November 3, 2017"
output: html_document
---

## Read file from github

```{r}
tanz_m <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-imputations.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Data Prep

```{r}

# convert categoricals to factors
tanz_m$basin <- factor(tanz_m$basin)
tanz_m$region <- factor(tanz_m$region)
tanz_m$region_code <- factor(tanz_m$region_code)
tanz_m$district_code <- factor(tanz_m$district_code)
tanz_m$public_meeting_new <- factor(tanz_m$public_meeting_new)
tanz_m$scheme_management <- factor(tanz_m$scheme_management)
tanz_m$permit_new <- factor(tanz_m$permit_new)
tanz_m$extraction_type <- factor(tanz_m$extraction_type)
tanz_m$extraction_type_class <- factor(tanz_m$extraction_type_class)
tanz_m$management <- factor(tanz_m$management)

tanz_m$payment <- factor(tanz_m$payment)
tanz_m$water_quality <- factor(tanz_m$water_quality)

tanz_m$quantity <- factor(tanz_m$quantity)
tanz_m$source <- factor(tanz_m$source)

tanz_m$waterpoint_type <- factor(tanz_m$waterpoint_type)


# convert labels to integers
tanz_m$label[which(tanz_m$label == 'functional')] <- 0
tanz_m$label[which(tanz_m$label == 'functional needs repair')] <- 1
tanz_m$label[which(tanz_m$label == 'non functional')] <- 2

tanz_m$label <- factor(tanz_m$label)

library(caret)

# split the Dataset in to train and validation subsets
set.seed(3033)
intrain <- createDataPartition(y = tanz_m$label, p= 0.7, list = FALSE)
training <- tanz_m[intrain,]
testing <- tanz_m[-intrain,]

```

Data is now prepped for model building.

_____

## Build a boosted classification tree model

79.65% accuracy. 65.31% accurate for "functional needs repair"

```{r}
library(C50)

bt_mod <- C5.0(label ~                        
quantity + waterpoint_type + extraction_type + pump_age + management +
                      region_code + district_code + water_quality + payment +
                      basin + source + scheme_management + latitude + longitude +
                      public_meeting_new + permit_new + amount_tsh_new + 
                      population_new +
                      installer_0 +
                      installer_cen +
                      installer_ces +
                      installer_com +
                      installer_dan +
                      installer_dis +
                      installer_dwe +
                      installer_fin +
                      installer_gov +
                      installer_hes +
                      installer_kkk +
                      installer_other +
                      installer_rwe +
                      installer_tcr +
                      installer_unk +
                      installer_wor +
                      funder_dan +
                      funder_dhv +
                      funder_dis +
                      funder_dws +
                      funder_fin +
                      funder_gov +
                      funder_hes +
                      funder_kkk +
                      funder_other +
                      funder_pri +
                      funder_rws +
                      funder_tas +
                      funder_uni +
                      funder_unk +
                      funder_wat +
                      funder_wor +
                      lga_freq_bin_1 +
                      lga_freq_bin_10 +
                      lga_freq_bin_11 +
                      lga_freq_bin_2 +
                      lga_freq_bin_20 +
                      lga_freq_bin_3 +
                      lga_freq_bin_4 +
                      lga_freq_bin_5 +
                      lga_freq_bin_6 +
                      lga_freq_bin_7 +
                      lga_freq_bin_8 +
                      lga_freq_bin_9 +
                      ward_freq_bin_1 +
                      ward_freq_bin_10 +
                      ward_freq_bin_11 +
                      ward_freq_bin_12 +
                      ward_freq_bin_13 +
                      ward_freq_bin_14 +
                      ward_freq_bin_15 +
                      ward_freq_bin_16 +
                      ward_freq_bin_17 +
                      ward_freq_bin_2 +
                      ward_freq_bin_20 +
                      ward_freq_bin_3 +
                      ward_freq_bin_4 +
                      ward_freq_bin_5 +
                      ward_freq_bin_6 +
                      ward_freq_bin_7 +
                      ward_freq_bin_8 +
                      ward_freq_bin_9 +
                   #   subvillage_freq_bin_1 +
                  #    subvillage_freq_bin_10 +
                  #    subvillage_freq_bin_11 +
                  #    subvillage_freq_bin_15 +
                  #    subvillage_freq_bin_2 +
                  #    subvillage_freq_bin_20 +
                  #    subvillage_freq_bin_3 +
                  #    subvillage_freq_bin_4 +
                  ##    subvillage_freq_bin_5 +
                  #    subvillage_freq_bin_6 +
                  #    subvillage_freq_bin_7 +
                  #    subvillage_freq_bin_8 +
                      gps_height_new,
                 trials = 20, # rules = TRUE,
            #     data = training)
     data = tanz_m)

# summary(bt_mod)

# remove the response variable from the validation subset
# bt_testing <- subset(testing, select = -c(label))
bt_testing <- subset(tanz_m, select = -c(label))

# make predictions for the validation
bt1 <- predict(bt_mod, bt_testing, type = "class")

# confusion matrix - validation subset
# bt_tab <- table(bt1, testing$label)
bt_tab <- table(bt1, tanz_m$label)
bt_tab



# display: off-diag numbers are mis-classifications
confusionMatrix(bt_tab)

```

## Confusion Matrix and Statistics

    
Confusion Matrix and Statistics

   
bt1     0     1     2
  0 31169  1378  2245
  1   216  2630   109
  2   874   309 20470

Overall Statistics
                                          
               Accuracy : 0.9136          
                 95% CI : (0.9113, 0.9159)
    No Information Rate : 0.5431          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.8395          
 Mcnemar's Test P-Value : < 2.2e-16       

Statistics by Class:

                     Class: 0 Class: 1 Class: 2
Sensitivity            0.9662  0.60922   0.8969
Specificity            0.8665  0.99410   0.9677
Pos Pred Value         0.8959  0.89002   0.9454
Neg Pred Value         0.9557  0.97011   0.9376
Prevalence             0.5431  0.07268   0.3842
Detection Rate         0.5247  0.04428   0.3446
Detection Prevalence   0.5857  0.04975   0.3645
Balanced Accuracy      0.9164  0.80166   0.9323

## Check variable importance:

```{r}
importance <- varImp(bt_mod)

importance <- importance[order(importance$Overall,decreasing = TRUE), ,drop = FALSE]

print(importance)

# memory cleanup
rm(bt1, bt_testing, bt_tab, importance)

```

                   Overall
quantity            100.00
extraction_type     100.00
lga_freq_bin_10     100.00
ward_freq_bin_16    100.00
ward_freq_bin_20    100.00
longitude            99.99
waterpoint_type      99.97
management           99.84
ward_freq_bin_11     99.28
pump_age             99.24
source               98.99
amount_tsh_new       98.90
region_code          98.62
water_quality        98.62
latitude             97.53
payment              97.24
district_code        96.90
scheme_management    95.25
funder_dws           92.77
gps_height_new       92.30
population_new       92.21
lga_freq_bin_20      92.04
installer_ces        89.43
ward_freq_bin_17     89.43
ward_freq_bin_8      88.22
installer_rwe        86.90
basin                86.71
ward_freq_bin_10     83.44
ward_freq_bin_7      82.98
funder_pri           82.54
installer_wor        81.31
installer_0          81.22
ward_freq_bin_12     81.20
funder_rws           80.87
installer_com        80.66
installer_cen        80.03
installer_gov        79.21
lga_freq_bin_1       78.87
public_meeting_new   76.77
funder_wor           75.98
funder_uni           75.71
installer_hes        75.25
ward_freq_bin_6      74.34
funder_hes           74.19
funder_dhv           71.43
lga_freq_bin_2       71.03
funder_dan           70.76
installer_fin        70.40
ward_freq_bin_1      69.92
ward_freq_bin_4      69.80
lga_freq_bin_11      69.56
funder_gov           68.41
ward_freq_bin_5      67.40
lga_freq_bin_3       67.36
funder_unk           67.09
funder_tas           65.18
ward_freq_bin_13     64.32
funder_wat           64.31
ward_freq_bin_14     63.98
lga_freq_bin_6       62.75
lga_freq_bin_8       62.66
ward_freq_bin_9      59.46
funder_dis           59.26
ward_freq_bin_2      58.27
installer_unk        57.53
ward_freq_bin_3      56.92
installer_tcr        56.65
installer_dwe        55.18
installer_dis        55.03
lga_freq_bin_7       53.77
lga_freq_bin_9       53.71
lga_freq_bin_4       53.34
lga_freq_bin_5       51.60
installer_dan        50.99
permit_new           50.13
funder_kkk           44.81
installer_kkk        39.89
installer_other      37.96
funder_other         36.33
funder_fin           31.87
ward_freq_bin_15      8.82

## Make predictions for submit data set

```{r}
# 
tanz_subm <- read.csv("C:/data/698/tanz-submit-imps.csv", header = TRUE, stringsAsFactors = FALSE)

# convert categoricals to factors
tanz_subm$basin <- factor(tanz_subm$basin)
tanz_subm$region <- factor(tanz_subm$region)
tanz_subm$region_code <- factor(tanz_subm$region_code)
tanz_subm$district_code <- factor(tanz_subm$district_code)
tanz_subm$public_meeting_new <- factor(tanz_subm$public_meeting_new)
tanz_subm$scheme_management <- factor(tanz_subm$scheme_management)

tanz_subm$permit_new <- factor(tanz_subm$permit_new)
tanz_subm$extraction_type <- factor(tanz_subm$extraction_type)
tanz_subm$extraction_type_class <- factor(tanz_subm$extraction_type_class)
tanz_subm$management <- factor(tanz_subm$management)

tanz_subm$payment <- factor(tanz_subm$payment)
tanz_subm$water_quality <- factor(tanz_subm$water_quality)

tanz_subm$quantity <- factor(tanz_subm$quantity)
tanz_subm$source <- factor(tanz_subm$source)

tanz_subm$waterpoint_type <- factor(tanz_subm$waterpoint_type)


# make predictions for the validation
btsub1 <- predict(bt_mod, tanz_subm)

# add predictions to submit dataframe
# add predictions to submit dataframe
tanz_subm$status_group <- as.character(btsub1)

# convert integer labels to corresponding text
tanz_subm$status_group[which(tanz_subm$status_group == '0')] <- 'functional'
tanz_subm$status_group[which(tanz_subm$status_group == '1')] <- 'functional needs repair'
tanz_subm$status_group[which(tanz_subm$status_group == '2')] <- 'non functional'

# write id + status_group cols to a file for submission

bt_preds <- cbind(tanz_subm$id, tanz_subm$status_group)
colnames(bt_preds) <- c("id", "status_group")


write.csv(bt_preds, file = "C:/data/698/bt_preds.csv", row.names = FALSE)

rm(bt_preds, btsub1)

```

