---
title: "tanz-mlr"
author: "James Topor"
date: "November 3, 2017"
output: html_document
---

## Read file from github

```{r}
tanz_m <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-imputations.csv", header = TRUE, stringsAsFactors = FALSE)
```


## Data Prep

```{r}

# convert categoricals to factors
tanz_m$basin <- factor(tanz_m$basin)
tanz_m$region <- factor(tanz_m$region)
tanz_m$region_code <- factor(tanz_m$region_code)
tanz_m$district_code <- factor(tanz_m$district_code)
tanz_m$public_meeting_new <- factor(tanz_m$public_meeting_new)
tanz_m$scheme_management <- factor(tanz_m$scheme_management)
tanz_m$permit_new <- factor(tanz_m$permit_new)
tanz_m$extraction_type <- factor(tanz_m$extraction_type)
tanz_m$extraction_type_class <- factor(tanz_m$extraction_type_class)
tanz_m$management <- factor(tanz_m$management)

tanz_m$payment <- factor(tanz_m$payment)
tanz_m$water_quality <- factor(tanz_m$water_quality)

tanz_m$quantity <- factor(tanz_m$quantity)
tanz_m$source <- factor(tanz_m$source)

tanz_m$waterpoint_type <- factor(tanz_m$waterpoint_type)


# convert labels to integers
tanz_m$label[which(tanz_m$label == 'functional')] <- 0
tanz_m$label[which(tanz_m$label == 'functional needs repair')] <- 1
tanz_m$label[which(tanz_m$label == 'non functional')] <- 2

tanz_m$label <- factor(tanz_m$label)

library(caret)

# split the Dataset in to train and validation subsets
set.seed(3033)
intrain <- createDataPartition(y = tanz_m$label, p= 0.7, list = FALSE)
training <- tanz_m[intrain,]
testing <- tanz_m[-intrain,]

```

Data is now prepped for model building.

_____

## Build a multinomial logistic regression model

Construct a model by adding one variable at a time in descending order of predictive value as determined via simple MLR models created using each individual independent variable as the sole predictor.



```{r}
library(nnet) # for multinom logistic regression

mlr1 <- multinom(label ~                        
quantity + waterpoint_type + extraction_type + pump_age + management +
                      region_code + district_code + water_quality + payment +
                      basin + source + scheme_management + latitude + longitude +
                      public_meeting_new + permit_new + amount_tsh_new + 
                      population_new +
                      installer_0 +
                      installer_cen +
                      installer_ces +
                      installer_com +
                      installer_dan +
                      installer_dis +
                      installer_dwe +
                      installer_fin +
                      installer_gov +
                      installer_hes +
                      installer_kkk +
                      installer_other +
                      installer_rwe +
                      installer_tcr +
                      installer_unk +
                      installer_wor +
                      funder_dan +
                      funder_dhv +
                      funder_dis +
                      funder_dws +
                      funder_fin +
                      funder_gov +
                      funder_hes +
                      funder_kkk +
                      funder_other +
                      funder_pri +
                      funder_rws +
                      funder_tas +
                      funder_uni +
                      funder_unk +
                      funder_wat +
                      funder_wor +
                      lga_freq_bin_1 +
                      lga_freq_bin_10 +
                      lga_freq_bin_11 +
                      lga_freq_bin_2 +
                      lga_freq_bin_20 +
                      lga_freq_bin_3 +
                      lga_freq_bin_4 +
                      lga_freq_bin_5 +
                      lga_freq_bin_6 +
                      lga_freq_bin_7 +
                      lga_freq_bin_8 +
                      lga_freq_bin_9 +
                      ward_freq_bin_1 +
                      ward_freq_bin_10 +
                      ward_freq_bin_11 +
                      ward_freq_bin_12 +
                      ward_freq_bin_13 +
                      ward_freq_bin_14 +
                      ward_freq_bin_15 +
                      ward_freq_bin_16 +
                      ward_freq_bin_17 +
                      ward_freq_bin_2 +
                      ward_freq_bin_20 +
                      ward_freq_bin_3 +
                      ward_freq_bin_4 +
                      ward_freq_bin_5 +
                      ward_freq_bin_6 +
                      ward_freq_bin_7 +
                      ward_freq_bin_8 +
                      ward_freq_bin_9,
  
                     # subvillage_freq_bin_1 +
                    #  subvillage_freq_bin_10 +
                     # subvillage_freq_bin_11 +
                    #  subvillage_freq_bin_15 +
                    #  subvillage_freq_bin_2 +
                    #  subvillage_freq_bin_20 +
                    #  subvillage_freq_bin_3 +
                    #  subvillage_freq_bin_4 +
                    #  subvillage_freq_bin_5 +
                    #  subvillage_freq_bin_6 +
                    #  subvillage_freq_bin_7 +
                    #  subvillage_freq_bin_8,

                 data = tanz_m, maxit = 200)

# confusion matrix - training subset
cm1 <- table(predict(mlr1, tanz_m), tanz_m$label)
# display: off-diag numbers are mis-classifications
cm1

# calculate percentage of accurate classifications
sum(diag(cm1))/sum(cm1)

mlr1$AIC

#summary(mlr1)

#pp <- fitted(mlr1)
#head(pp, n = 20)

# -----------------------



# memory cleanup
rm(cm1, cm2)

# use step() function to check for variables that might be discarded
# mmodi <- step(mlr1)
```


## Check variable importance:

```{r}
importance <- varImp(mlr1)

importance <- importance[order(importance$Overall,decreasing = TRUE), ,drop = FALSE]

print(importance)

# memory cleanup
rm(gbm1, gbm_testing, gbm_tab, gbm_mod)

```

_____ 

Now do cross validation using model constructed above: 75.08% cross validated accuracy. 

```{r, eval = FALSE}

trctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
set.seed(333)

mlr_fit <- train(label ~                        
quantity + waterpoint_type + extraction_type + pump_age + management +
                      region_code + district_code + water_quality + payment +
                      basin + source + scheme_management + latitude + longitude +
                      public_meeting_new + permit_new + amount_tsh_new + 
                      population_new +
                      installer_0 +
                      installer_cen +
                      installer_ces +
                      installer_com +
                      installer_dan +
                      installer_dis +
                      installer_dwe +
                      installer_fin +
                      installer_gov +
                      installer_hes +
                      installer_kkk +
                      installer_other +
                      installer_rwe +
                      installer_tcr +
                      installer_unk +
                      installer_wor +
                      funder_dan +
                      funder_dhv +
                      funder_dis +
                      funder_dws +
                      funder_fin +
                      funder_gov +
                      funder_hes +
                      funder_kkk +
                      funder_other +
                      funder_pri +
                      funder_rws +
                      funder_tas +
                      funder_uni +
                      funder_unk +
                      funder_wat +
                      funder_wor +
                      lga_freq_bin_1 +
                      lga_freq_bin_10 +
                      lga_freq_bin_11 +
                      lga_freq_bin_2 +
                      lga_freq_bin_20 +
                      lga_freq_bin_3 +
                      lga_freq_bin_4 +
                      lga_freq_bin_5 +
                      lga_freq_bin_6 +
                      lga_freq_bin_7 +
                      lga_freq_bin_8 +
                      lga_freq_bin_9 +
                      ward_freq_bin_1 +
                      ward_freq_bin_10 +
                      ward_freq_bin_11 +
                      ward_freq_bin_12 +
                      ward_freq_bin_13 +
                      ward_freq_bin_14 +
                      ward_freq_bin_15 +
                      ward_freq_bin_16 +
                      ward_freq_bin_17 +
                      ward_freq_bin_2 +
                      ward_freq_bin_20 +
                      ward_freq_bin_3 +
                      ward_freq_bin_4 +
                      ward_freq_bin_5 +
                      ward_freq_bin_6 +
                      ward_freq_bin_7 +
                      ward_freq_bin_8 +
                      ward_freq_bin_9,

                      data = training, method = "multinom",
                      maxit = 200,
                      trControl=trctrl,
                      preProcess = c("center", "scale"),
                      tuneLength = 10
        )

# check results
mlr_fit
````

Penalized Multinomial Regression 

41581 samples
   80 predictor
    3 classes: '0', '1', '2' 

Pre-processing: centered (194), scaled (194) 
Resampling: Cross-Validated (5 fold, repeated 2 times) 
Summary of sample sizes: 33266, 33264, 33265, 33264, 33265, 33265, ... 
Resampling results across tuning parameters:

  decay         Accuracy   Kappa    
  0.0000000000  0.7509677  0.5156793
  0.0001000000  0.7510158  0.5157656
  0.0002371374  0.7509798  0.5156985
  0.0005623413  0.7509557  0.5156735
  0.0013335214  0.7508475  0.5154472
  0.0031622777  0.7509317  0.5156123
  0.0074989421  0.7511721  0.5161091
  0.0177827941  0.7509798  0.5157172
  0.0421696503  0.7510038  0.5158278
  0.1000000000  0.7511120  0.5160545

Accuracy was used to select the optimal model using  the largest value.
The final value used for the model was decay = 0.007498942.

```{r, eval = FALSE}
# get predictions
test_pred <- predict(mlr_fit, newdata = testing)
# test_pred

# check confusion matrix
confusionMatrix(test_pred, testing$label )
```
Confusion Matrix and Statistics

          Reference
Prediction    0    1    2
         0 8716  895 2279
         1   67  147   52
         2  894  253 4516

Overall Statistics
                                          
               Accuracy : 0.7508          
                 95% CI : (0.7444, 0.7572)
    No Information Rate : 0.5431          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.5156          
 Mcnemar's Test P-Value : < 2.2e-16       

Statistics by Class:

                     Class: 0 Class: 1 Class: 2
Sensitivity            0.9007  0.11351   0.6596
Specificity            0.6102  0.99280   0.8955
Pos Pred Value         0.7331  0.55263   0.7975
Neg Pred Value         0.8379  0.93460   0.8082
Prevalence             0.5431  0.07268   0.3843
Detection Rate         0.4891  0.00825   0.2534
Detection Prevalence   0.6673  0.01493   0.3178
Balanced Accuracy      0.7554  0.55316   0.7775


## Make predictions for submit data set

```{r}
# 
tanz_subm <- read.csv("https://raw.githubusercontent.com/jtopor/DDO-TNZ/master/tanz-submit-imps.csv", header = TRUE, stringsAsFactors = FALSE)

# convert categoricals to factors
tanz_subm$basin <- factor(tanz_subm$basin)
tanz_subm$region <- factor(tanz_subm$region)
tanz_subm$region_code <- factor(tanz_subm$region_code)
tanz_subm$district_code <- factor(tanz_subm$district_code)
tanz_subm$public_meeting_new <- factor(tanz_subm$public_meeting_new)
tanz_subm$scheme_management <- factor(tanz_subm$scheme_management)

tanz_subm$permit_new <- factor(tanz_subm$permit_new)
tanz_subm$extraction_type <- factor(tanz_subm$extraction_type)
tanz_subm$extraction_type_class <- factor(tanz_subm$extraction_type_class)
tanz_subm$management <- factor(tanz_subm$management)

tanz_subm$payment <- factor(tanz_subm$payment)
tanz_subm$water_quality <- factor(tanz_subm$water_quality)

tanz_subm$quantity <- factor(tanz_subm$quantity)
tanz_subm$source <- factor(tanz_subm$source)

tanz_subm$waterpoint_type <- factor(tanz_subm$waterpoint_type)


# make predictions for the validation
mlrsub1 <- predict(mlr1, tanz_subm)

# add predictions to submit dataframe
# add predictions to submit dataframe
tanz_subm$status_group <- as.character(mlrsub1)

# convert integer labels to corresponding text
tanz_subm$status_group[which(tanz_subm$status_group == '0')] <- 'functional'
tanz_subm$status_group[which(tanz_subm$status_group == '1')] <- 'functional needs repair'
tanz_subm$status_group[which(tanz_subm$status_group == '2')] <- 'non functional'

# write id + status_group cols to a file for submission

mlr_preds <- cbind(tanz_subm$id, tanz_subm$status_group)
colnames(mlr_preds) <- c("id", "status_group")


write.csv(mlr_preds, file = "C:/data/698/mlr_preds.csv", row.names = FALSE)

rm(mlr_preds, mlrsub1)

```
